---
title: "Lab 13LC"
author: "Lucas Collignon"
format:
  html:
    toc: true
    toc_float: true
    embed-resources: true
    code-fold: true
    theme: darkly
execute:
  echo: true
  freeze: auto
  warning: false
  message: false
---

```{r}
.libPaths(c("/home/lcollignon_umass_edu/R/x86_64-pc-linux-gnu-library/4.5", "/usr/local/lib/R/site-library", "/usr/local/lib/R/library"))
```

# Community Composition & Diversity Tutorial – NEON Soil Chemistry

```{r}
# Load packages
library(tidyverse)
library(vegan)
library(janitor)
library(ggplot2)
```

```{r}
soil <- read_csv("NEON_soilMAGs_soilChem.csv") %>% clean_names()

glimpse(soil)
```
```{r}
soil <- soil %>%
  separate(genomics_sample_id,
           into = c("site", "subplot", "layer", "date", "comp"),
           sep = "-",
           remove = FALSE,
           fill = "right")
```
## Environmental Profiling
**Environmental gradients strongly influence microbial community structure. Profiling helps identify which factors may drive differences among sites, explains differences between levels in different environments**
```{r}
env_summary <- soil %>%
  group_by(site) %>%
  summarise(
    mean_temp = mean(soil_temp, na.rm = TRUE),
    mean_moisture = mean(soil_moisture, na.rm = TRUE),
    mean_water_ph = mean(soil_in_waterp_h, na.rm = TRUE),
    mean_cacl_ph = mean(soil_in_ca_clp_h, na.rm = TRUE)
  )
env_summary
```
## By site

```{r}
env_long <- env_summary %>%
  pivot_longer(-site, names_to = "variable", values_to = "value")

ggplot(env_long, aes(x = site, y = value, fill = variable)) +
  geom_col(position = "dodge") +
  theme_bw() +
  labs(title = "Environmental Composition Across NEON Sites",
       x = "Site", y = "Mean Value")
```
## Alpha Diversity of Environmental Variables

```{r}
env_matrix <- soil %>%
  group_by(site) %>%
  summarise(
    temp = mean(soil_temp, na.rm = TRUE),
    moisture = mean(soil_moisture, na.rm = TRUE),
    water_ph = mean(soil_in_waterp_h, na.rm = TRUE),
    cacl_ph = mean(soil_in_ca_clp_h, na.rm = TRUE)
  ) %>%
  column_to_rownames("site")
```

## Shannon index, Simpson Index, richness
**The shannon index explains the balance between richness and evenness. A high shannon says that all environmental factors had equal contribution, low shannon says one factor has a large influence**

**richness is how many variables are recorded per sample**

**Simpson index is somewhat opposite to Shannon, if there is a low simpson one factor is dominating and the reason for change, opposite for high simpson**
```{r}
shannon <- diversity(env_matrix, index = "shannon")
simpson <- diversity(env_matrix, index = "simpson")
richness <- specnumber(env_matrix)

alpha_df <- tibble(
  site = rownames(env_matrix),
  shannon = shannon,
  simpson = simpson,
  richness = richness
)
alpha_df
```
## Shannon plot

```{r}
ggplot(alpha_df, aes(x = site, y = shannon)) +
  geom_col(fill = "steelblue") +
  theme_bw() +
  labs(title = "Shannon Diversity of Environmental Variables",
       x = "Site", y = "Shannon Index")
```
## bray curtis

**this is a beta diversity measure. closer to 0 is closer to identical environments, 1 is different environments**
```{r}
bc_dist <- vegdist(env_matrix, method = "bray")
bc_dist
```

## Principal component analysis

**PCA identifies major gradients in soil chemistry**
```{r}
pca <- prcomp(env_matrix, scale. = TRUE)

pca_df <- as.data.frame(pca$x)
pca_df$site <- rownames(env_matrix)

ggplot(pca_df, aes(x = PC1, y = PC2, label = site)) +
  geom_point(size = 4, color = "darkblue") +
  geom_text(vjust = -0.5) +
  theme_bw() +
  labs(title = "PCA of Soil Environmental Variables",
       x = paste0("PC1 (", round(summary(pca)$importance[2,1] * 100, 1), "%)"),
       y = paste0("PC2 (", round(summary(pca)$importance[2,2] * 100, 1), "%)"))
```

## Non metric multidimensional scaling

**another form of environmental analysis, closer points mean more similar environments**

```{r}
# Prepare numeric matrix for NMDS
env_matrix_samples <- soil %>%
  select(genomics_sample_id,
         soil_temp, soil_moisture, soil_in_waterp_h, soil_in_ca_clp_h) %>%
  drop_na()

env_numeric <- env_matrix_samples %>%
  select(-genomics_sample_id)

# Run NMDS
library(vegan)
set.seed(123)
nmds <- metaMDS(env_numeric, distance = "bray", k = 2, trymax = 100)

# Extract site scores only
site_scores <- scores(nmds, display = "sites")

# Convert to dataframe
nmds_df <- as.data.frame(site_scores)

# Add sample IDs
nmds_df$sample_id <- env_matrix_samples$genomics_sample_id

# Plot NMDS
ggplot(nmds_df, aes(x = NMDS1, y = NMDS2, label = sample_id)) +
  geom_point(size = 4, color = "darkred") +
  geom_text(vjust = -0.5) +
  theme_bw() +
  labs(title = "NMDS of Soil Environmental Variables",
       x = "NMDS1",
       y = "NMDS2")
```





